---
title: "A Guide for Instructors"
output:
    bookdown::html_document2:
      base_format: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SppDistMonProj}
  %\VignetteEngine{bookdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

The SppDistMonProj package was developed to be used as an aid for teaching data management and workflow concepts using R. This vignette provides an introduction of the SppDistMonProj R package for a course instructor. It provides step-by-step instructions on how to use the package to teach applied data management concepts.


## Overview

### The Example Survey

This package uses a simulated survey of species occurrence to demonstate how R can be used to develop a data workflow that includes many steps of a data life cycle including importing, processing, analyzing and reporting on data. The objective of the survey is to estimate the occupancy of a hypothetical species within a define square sampling frame (a "refuge"). Occupancy is a function of environmental covariates. To do this, package functions simulate the spatial distribution of environmental covariates in the sampling frame. Based on the distribution of these covariates, a package function simulates the distribution of the target species of interest. Sampling data can then either be simulated or generated as an in-class exercise using ArcGIS Survey123. These data are summarized in a quality control RMarkdown report and in an analysis Rmarkdown report.   

### Using SppDistMonProj for Teaching

The first step is generate data, either through simulation (detailed in [Simulating data][Simulating data]) or by having the student "collect" data using an iPad configured with a customized ArcGIS Survey123 app.  Next, the instructor introduces students to a basic data workflow in R that starts with developing scripts to import, process, analyze and report on data, bundling these scripts into customized functions, and finally assembling and documenting these customized functions in an R package. 


## Step 1: Acquiring data

### Option A. Simulating data

If you do not have any data in hand, the first step is to generate some data through simulation. `SppDistMonProj::create.raster()` generates a raster (.asc) of covariate values.  `SppDistMonProj::one.step()` simulates, formats and saves survey site `site.data` and observation data `occu.data`. Observation data are outputted as `unmarked::occu.data` formatted dataframes and as raw data output would be generated from ArcGIS Survey123 app (`s123.data`). The following scripts show the steps needed to create these datasets.

```{r, message=FALSE}
# Load required package
library(raster)
library(SppDistMonProj)
```

Create a project directory
```{r}
SppDistMonProj::create.dir(proj.name = "course_survey")
```

Create and save a raster of habitat covariates

```{r}
make.rast <- TRUE  # Set as true for the first time running

# Create spatial covariate values for sites (i.e., raster cells)
if(make.rast){
    rast <- create.raster()
    habcov <- data.frame(id = as.factor(1:ncell(rast)), 
                         rasterToPoints(rast))
    
    # Save RasterLayer as ESRI ascii format
    writeRaster(rast, 
                file = "./resources/data/geodata/habcov.asc", 
                format = "ascii", 
                overwrite = TRUE)
} else {
    rast <- raster("./resources/data/geodata/habcov.asc")
    habcov <- data.frame(id = as.factor(1:ncell(rast)), 
                         rasterToPoints(rast))
}
```


Simulate, format and save a list of survey site and observation data
```{r, eval=FALSE}
## Set simulation parameters
years <- 2011:2019
nsites <- 20
nocc <- 5
p <- data.frame(label = c("good","poor"), p = c(0.7,0.3))
write.data <- TRUE
write.images <- FALSE
return.data <- TRUE
ngood <- 350
npoor <- 500
cex <- 1.75
showcoords <- FALSE

# Simulate, format and save a list of survey site and observation data for as many 
# years as specified above (9 years in this example)
data.list <- mapply(one.step, year = years,
                    MoreArgs = list(habcov = habcov, 
                                    nsites = nsites, 
                                    nocc = nocc, 
                                    p = p, 
                                    write.data = write.data,
                                    write.images = write.images, 
                                    ngood = ngood, 
                                    npoor = npoor,
                                    cex = 1.75, 
                                    showcoords = showcoords, 
                                    return.data = return.data),
                    SIMPLIFY = FALSE)
```

### Option B. Collecting data using ArcGIS Survey123

A second option is to have students collect "data" as an in-class exercise using ArcGIS Survey123. The example Survey123 project is located [here](). Contact [Jared Laufenberg](mailto:jared_laufenberg@fws.gov) for access.

Need content here


### Annual sample site selection

```{r}
# Load packages
library(raster)
```

```{r}
# Load spatial data
habcov.raster <- raster("./resources/data/geodata/habcov.asc")
ncells <- nrow(habcov.raster) * ncol(habcov.raster)
habcov <- data.frame(id = as.factor(1:ncells), rasterToPoints(habcov.raster))
```

```{r}
# Randomly select sites to be surveyed in specified year
year <- 2020
nsites <- 20
set.seed(year) # ensures same sample drawn each time code is run for a given year
surveysites.i <- habcov[sample(1:ncells, nsites),]
```

```{r}
# Randomly select sites to be surveyed in specified year
year <- 2020
nsites <- 20
set.seed(year) # ensures same sample drawn each time code is run for a given year
surveysites.i <- habcov[sample(1:ncells, nsites),]
```

```{r}
# Plot survey sites
plot(habcov.raster, xlab = "X", ylab = "Y", asp = 1, main = "Blue crosses = selected sites")
points(surveysites.i$x, surveysites.i$y, pch = "+", col = "blue", cex = 2)
```


## Step 3. Process the data

The following describes the teaching steps that demonstrate the general progression from developing scripts, bundling scripts into custom functions, and then finally bundling and documenting custom functions as an R package.

### Load packages

The first step is to load the required packages

```{r}
library(unmarked)
library(raster)
```

### Read raw Survey123 data files and create a data object

For this exampleNext, we need to read in the raw occurrence and sample site data. If they are all saved in the same folder we can first list these files:

```{r}
# observation data file names
(o.files <- list.files("../data/raw", 
                       pattern = "occurrence", 
                       full.names = TRUE))

# site data file names
(s.files <- list.files("../data/raw", 
                       pattern = "sample_site", 
                       full.names = TRUE))
```


And then determine how many replicate surveys occurred:

```{r}
(nfiles <- length(o.files))
```

Next we can create data object for survey years using a "for loop":

```{r}
years <- rep(NA, nfiles)
 
for(i in 1:nfiles){
    ind = gregexpr("_", o.files[i])[[1]][2]
    years[i] <- as.numeric(substr(o.files[i], ind+1, ind+4))
}

years
```


```{r}
## create data object for survey years using custom function and sapply function
extract.years <- function(filename){
    ind = gregexpr("_", filename)[[1]][2]
    substr(filename, ind + 1, ind + 4)
}
(years <- as.numeric(sapply(o.files, extract.years)))

(nsfiles <- length(s.files))
```


## Quality control

Load packages
```{r}
library(unmarked)
library(raster)
```

Load the processed data
```{r}
year <- 2018
load(file = paste0("../data/final/species_occurrence_data_", year, ".gzip"))
str(data.list, max.level=1)
str(data.list, max.level=2)
```

### Perform data quality checks with functions from `unmarked` package

```{r}
str(data.list[[1]]$unmarked.data)

summary(data.list[[1]]$unmarked.data)

plot(data.list[[1]]$unmarked.data)

numSites(data.list[[1]]$unmarked.data)

numY(data.list[[1]]$unmarked.data)

obsNum(data.list[[1]]$unmarked.data)

show(data.list[[1]]$unmarked.data)

obsCovs(data.list[[1]]$unmarked.data)

siteCovs(data.list[[1]]$unmarked.data)
```

### Additional data quality checks

Create years data object
```{r}
years <- sapply(data.list, function(x)x$year)
```

Calculate naive site occupancy status for sampled sites
```{r}
apply(slot(data.list[[1]]$unmarked.data, "y"), 1, max)
```

Use custom function and more apply functions
```{r}
naive.occu <- function(x){apply(slot(x$unmarked.data, "y"), 1, max)}
naiveoccu <- lapply(data.list, naive.occu)
str(naiveoccu)
```

Calculate annual naive occupancy probability for samples sites
```{r}
naiveoccu.byyear <- sapply(naiveoccu, mean)
names(naiveoccu.byyear) <- years
naiveoccu.byyear
```

Create and save figure for annual naive occupancy probabilities
```{r}
plot(years, naiveoccu.byyear, 
     ylim = c(0,1), 
     type = "b", 
     pch = 16, 
     frame.plot = FALSE, 
     axes  =  FALSE,
     ylab = "Naive proportion of sites occupied", 
     xlab = "Year")
axis(2)
axis(1, at = years)
```

Tabulate number of surveys conducted by each observer for each site
```{r}
table(data.list[[1]]$occdf[,c("id","observer")])

annual.tables <- function(x,year){write.csv(table(x$occdf[,c("id","observer")]),
                                            paste0("../output/tables/observer_table_", 
                                                   year, ".csv"))}
mapply(annual.tables, x = data.list, year = years)
```

Plot map of naive occupancy
```{r}
habcov.raster <- raster("../resources/data/geodata/habcov.asc")
sitecovs <- data.frame(habcov = values(habcov.raster))
nsites <- length(habcov.raster)
rasterpts <- data.frame(rasterToPoints(habcov.raster))
ss <- as.numeric(names(naiveoccu[[1]]))

plot(habcov.raster, xlab = "X", ylab = "Y", asp = 1)
points(rasterpts$x[ss], 
       rasterpts$y[ss], 
       pch = c(21,16)[naiveoccu[[1]] + 1],
       col = "blue", 
       bg = c(NA,"blue")[naiveoccu[[1]] + 1], 
       cex = 2)

plot.annual.naiveoccu <- function(habcov.raster,naiveoccu,year){
    png(paste0("../output/figures/naiveoccu_map_",year,".png"), 
        width = 6.5, 
        height = 6.5, 
        units = "in", 
        res = 96)
    rasterpts <- data.frame(rasterToPoints(habcov.raster))
    ss <- as.numeric(names(naiveoccu))
    plot(habcov.raster, xlab = "X", ylab = "Y", asp = 1)
    points(rasterpts$x[ss], 
           rasterpts$y[ss], 
           pch = c(21,16)[naiveoccu + 1],
           col = "blue", 
           bg = c(NA,"blue")[naiveoccu + 1], 
           cex = 2)
    dev.off()
}

mapply(plot.annual.naiveoccu, 
       naiveoccu = naiveoccu, 
       year = years, 
       MoreArgs = list(habcov.raster = habcov.raster))
```

Check sample distribution of habitat covaraite values
```{r}
hist(data.list[[1]]$sitedf$habcov, breaks = seq(-3,3,0.2), ylim = c(0,10),
     main = "", xlab = "Habitat covariate")
```

### Class exercise: create a function similar to plot.annual.naiveoccu and use it to create & save a histogram for each year

```{r}
plot.sample.habcov.hist <- function(){}

plot.sample.habcov.hist <- function(x, year){
  png(paste0("../output/figures/sample_habcov_hist_", year, ".png"), 
      width = 6.5, height = 6.5, units = "in", res = 96)
  hist(x$sitedf$habcov, breaks = seq(-3,3,0.2), ylim = c(0,10),
      main = "", xlab = "Habitat covariate")
  dev.off()
}

mapply(plot.sample.habcov.hist, x = data.list, year = years)

```

### Check for same years in occurrence and site data file names 

Updated custom function
```{r}
extract.years <- function(filename){
    if(grepl("occurrence", filename)) ind = gregexpr("_", filename)[[1]][2]
    if(grepl("sample_site", filename)) ind = gregexpr("_", filename)[[1]][3]
    substr(filename, ind + 1, ind + 4)
}
```

Read file names, extract years for each data type, and compare years
```{r}
o.files <- list.files("../data/raw", 
                      pattern = "occurrence", 
                      full.names = TRUE)
s.files <- list.files("../data/raw", 
                      pattern = "sample_site", 
                      full.names = TRUE)
(o.years <- as.numeric(sapply(o.files, extract.years)))
(s.years <- as.numeric(sapply(s.files, extract.years)))
(miss.o <- !(s.years %in% o.years))
(miss.s <- !(o.years %in% s.years))
```


## Data analysis

Load packages
```{r}
library(raster)
library(unmarked)
```

Load the data
```{r}
year <- 2018
load(paste0("../data/final/species_occurrence_data_", 
            year, 
            ".gzip"))
years <- unlist(lapply(data.list, function(x)x$year))
nyears <- length(years)
habcov.raster <- raster("../resources/data/geodata/habcov.asc")
sitecovs <- data.frame(habcov = values(habcov.raster))
nsites <- length(habcov.raster)
```

Fit models and estimate parameters
```{r}
# specify candidate model set
models <- c(Null = as.formula("~ 1 ~ 1"), 
            DetCov = as.formula("~ viscov ~ 1"),
            OccuCov = as.formula("~ 1 ~ habcov"), 
            Global = as.formula("~ viscov ~ habcov"))
nmods <- length(models)
modnames <- c("Null","DetCov","OccuCov","Global")

# fit all candidate models to each year of data
allfits <- lapply(data.list, function(data)fitList(fits = lapply(models,function(x,data)occu(x,data),
                                                                 data = data$unmarked.data)))
```

Process the results
```{r}
modeltables <- lapply(allfits, function(x)as(modSel(x), "data.frame"))
modranks <- as.data.frame(lapply(modeltables, function(x)match(modnames, x$model)),
                          row.names = modnames, col.names = paste0("Y", years))
newdata1 <- data.frame(habcov = seq(-3,3,0.1) )
Epsi <- lapply(allfits, predict, type = "state", newdata = newdata1, appendData = TRUE)
newdata2 <- data.frame(viscov = factor(c("good","poor")))
Ep <- lapply(allfits, predict, type = "det", newdata = newdata2, appendData = TRUE)

```


```{r}
source("./functions/get_topmod.R")
(topmods <- mapply(get.topmod, allfits, modranks))
```

```{r}
source("./functions/get_betas.R")
(betas <- lapply(topmods, get.betas))
```

```{r}
source("./functions/plot_betas.R")
plot.betas(betas, years, modranks)
```


## Questions and Feedback

Contact [Jared Laufenberg](mailto: jared_laufenberg@fws.gov) or 
[McCrea Cobb](mailto: mccrea_cobb@fws.gov). 
